Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-05-09 WSSSPE2_report/WSSSPE2_review.tex
Section 3.1: Add URLs as footnotes.  Section 7: minor changes. Add a note.

Kaitlin Thaney is the Director of the Mozilla Science Lab (hereafter Mozilla).
Mozilla is a non-profit organization interested in openness, in news, in website
creation, and in Science, all taking advantage of the open web.

Thaney started noting the unfortunate fact that many current systems suffer the
unintended consequence of creating friction that hinders users, despite
designers' original purposes to do good. An example is the National Cancer
Institute's caBIG. A total of \$350 million was spent, including more than \$60
million for management. More than $70$ tools were created, but caBIG is still
seen as a failure\footnote{\url{http://tinyurl.com/maf6dz2}}. Those that had the
least investment were the most used; the most invested software were the least
utilized.

@@

Thaney spoke about Mozilla's work around code citation, through a collaboration
and prototype crafted between Mozilla, GitHub, figshare, and Zenodo. This work
was presented at a closed meeting in May 2014 at the National Institutes of
Health (NIH) around these issues, sparking a conversation from that meeting
around what a \emph{Software Discovery
Index}\footnote{\url{http://softwarediscoveryindex.org/report/}} might look
like. The meeting included a number of publishers, researchers, and those behind
major scientific software efforts such as Bioconductor, Galaxy, and
nanoHUB. 
%facilitates more efficient scientific research. SDI identifies scientific
%software by archiving and standardizing metadata for software and hence help
%connect both developers and users.
Ted Habermann in the audience commented that if the metadata is minimal, it
would be less onerous for data providers, but more burdensome for users---it
could be challenging to keep a balance between what have to be captured and what
would be ideal if we do not want to lose user engagement as in the case of the
old Harvard Dataverse, finding often only the first four fields of three pages
of metadata were filled out.

@@

We then discussed career paths for those supporting sustainable software. While
tenure track is not the only option for students, the challenges for those who
remain in academia can be large. Research scientists are entirely dependent on
soft money, which can be unpredictable. Postdocs and those who do pursue tenure
track positions need to publish and see no rewards in software development.
These challenges were all identified at WSSSPE1. What actions would we recommend
to improve things? Altmetrics and download statistics may slowly change the
system and improve a developer's ability to receive credits for time invested in
software development. NSF's recognition of data products (including software) as
well as publications is also helping.

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?'' One example demonstrates change
over time. In 2007, nanoHUB listed the academic reward structure as a problem in
an EDUCAUSE report where the authors note, ``In the future, nanoHUB researchers
are hoping to change the research culture. While they recognize that young
faculty members are unlikely to get tenure based on their nanoHUB contributions,
they hope to encourage faculty to think beyond their own research needs to
consider publishing tutorials and other content in their fields on the
nanoHUB site.'' Fast forward seven years later where, in a 2014 iSGTW article,
quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A former student of
mine published eight tools on nanoHUB, serving over 6,000 people with his tools.
He then joined a university as a professor and introduced nanoHUB. Use of the
gateway from that university skyrocketed; he used nanoHUB in existing classes,
created new classes, and infused it in his research. Ultimately, the professor's
department head associates his two-year rise to tenure with the notoriety and
innovation he gained through nanoHUB.''

The group also discussed ``attribution trees,'' an idea put forward by Dan Katz
and Arfon Smith where a chain of attributions can, for example, give credit to
those developing libraries and building blocks that other pieces of software
use. The group considered where to push the use of this, which journals? On
participant noted that Dryad works with the journals. If a paper is accepted,
the supporting data must be submitted. We discussed reproducibility as a
component of the journal review process and ``active papers,'' with immediate
links to the data and software.

>>>>

Kaitlin Thaney is the Director of the Mozilla Science Lab (hereafter Mozilla), which
is a non-profit organization interested in openness, news, website
creation, and Science, all taking advantage of the open web.

Thaney started noting the unfortunate fact that many current systems suffer the
unintended consequence of creating friction that hinders users, despite
designers' original purposes to do good. An example is the National Cancer
Institute's caBIG. A total of \$350 million was spent, including more than \$60
million for management. More than $70$ tools were created, but caBIG is still
seen as a failure\footnote{Report Blasts Problem-Plagued Cancer Research Grid,
\url{http://tinyurl.com/maf6dz2}}. Those that had the least investment were the
most used; the most invested software were the least utilized.

@@

Thaney spoke about Mozilla's work around code citation, through a collaboration
and prototype crafted between Mozilla, GitHub, figshare, and Zenodo. This work
was presented at a closed meeting in May 2014 at the National Institutes of
Health (NIH) around these issues, sparking a conversation from that meeting
around what a \emph{Software Discovery
Index}\footnote{Software Discovery Index, \url{http://softwarediscoveryindex.org/report/}} might look
like. The meeting included a number of publishers, researchers, and those behind
major scientific software efforts such as 
Bioconductor{\footnote{Bioconductor, \url{http://www.bioconductor.org}}, 
Galaxy\footnote{Galaxy, \url{http://galaxyproject.org}}, and
nanoHUB\footnote{nanoHUB, \url{https://nanohub.org}}. 
%facilitates more efficient scientific research. SDI identifies scientific
%software by archiving and standardizing metadata for software and hence help
%connect both developers and users.
Ted Habermann in the audience commented that if the metadata is minimal, it
would be less onerous for data providers, but more burdensome for users---it
could be challenging to keep a balance between what have to be captured and what
would be ideal if we do not want to lose user engagement as in the case of the
old Harvard Dataverse, finding often only the first four fields of three pages
of metadata were filled out.

@@

We then discussed career paths for those supporting sustainable software. While
tenure track is not the only option for graduate students, the challenges for those who
remain in academia can be large. Research scientists are entirely dependent on
soft money, which can be unpredictable. Postdocs and those who do pursue tenure
track positions need to publish and see no rewards in software development.
These challenges were all identified at WSSSPE1. What actions would we recommend
to improve things? Altmetrics and download statistics may slowly change the
system and improve a developer's ability to receive credits for time invested in
software development. NSF's recognition of scientific products including datasets, 
software, and publications is also helping.

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?'' One example demonstrates change
over time. In 2007, nanoHUB listed the academic reward structure as a problem in
an EDUCAUSE report where the authors note, ``In the future, nanoHUB researchers
are hoping to change the research culture. While they recognize that young
faculty members are unlikely to get tenure based on their nanoHUB contributions,
they hope to encourage faculty to think beyond their own research needs to
consider publishing tutorials and other content in their fields on the
nanoHUB site.'' Fast forward seven years later where, in a 2014 iSGTW article,
quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A former student of
mine published eight tools on nanoHUB, serving over 6,000 people with his tools.
He then joined a university as a professor and introduced nanoHUB. Use of the
gateway from that university skyrocketed; he used nanoHUB in existing classes,
created new classes, and infused it in his research. Ultimately, the professor's
department head attributed his two-year rise to tenure with the reputation and
innovation he gained through nanoHUB.''

The group also discussed ``attribution trees,'' an idea put forward by Dan Katz
and Arfon Smith where a chain of attributions can, for example, give credit to
those developing libraries and building blocks that other pieces of software
use. The group considered potential journals and medium to push the application
of this idea. One participant noted that Dryad\footnote{The Dryad Digital
Repository, \url{http://datadryad.org}.} works with journals. If a paper is
accepted, the supporting data must be submitted. We discussed reproducibility as
a component of the journal review process and ``active papers,'' with immediate
links to the data and software. \choinote{Connect the last two sentences to the preceeding materials.}
